Hybrid nonlinear system identification using neural networks

### 2020-11-24 systemID_NNs.tex
edits.:

in which $N_p$ is the number of past measurements and $\Delta$ is the sample
time between two consecutive measurements. There exists a $\phi(\cdot)$ and a
number of delay time steps $N_p$, such that the plant state at a given time can
be written as $x_p(k) = \phi(\mathbf{y}_{k-N_p:k}, \mathbf{u}_{k-N_p:k-1})$. For
linear plants, this function $\phi(\cdot)$ is known analytically in terms of the
plant dynamic matrices.

@@

\begin{align*}
  k_1 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k-N_p:k-1}, \mathbf{u}_{k-N_p:k-1}, u) \\
  x_g & := x_g + (\Delta/2)k_1 \\
  k_2 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g & := x_g + (\Delta/2)k_2 \\
  k_3 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g & := x_g + \Delta k_3 \\
  k_4 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g(k+\Delta) &= x_g(k) + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) 
\end{align*}
in which the sequence $\mathbf{y}_{k+0.5-N_p:k-0.5}$ is obtained by linear
interpolation.

The following optimization problem is solved to compute the weights in the
neural network
\begin{align*}
\underset{W_i, b_i}{\textnormal{min}} \sum_{k=0}^{N_s-1} 
\dfrac{1}{N_s}(\hat{y}(k) - y(k))^2
 \end{align*}
in which $W_i$, $b_i$ are the weights in each hidden layer of the NN, $N_s$ is
the number of training samples, $\hat{y}$ are the predictions of the augmented
grey-box + neural network model, $N_s$ is the number of training samples, and
$y$ are the measurements from the plant. Note that this optimization requires an
estimate of the initial grey-box state. We assume that before the open-loop
system identification experiment performed for data collection, the plant is
being operated at some steady-state such that input to output responses of the
plant and grey-box model are aligned with the use of an integrating disturbance
model. We can use the grey-box state at that steady state as the initial
condition for the above multi-step-ahead prediction-error optimization.

\subsection{Nonlinear reactor example}
We consider a nonlinear reactor with two reactions ($A \rightarrow B $ and  $3B
\rightleftharpoons C$), simulated using the following ordinary differential
equations (ODEs)
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - 3k_2C^3_B + 3k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C^3_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}

The grey-box model is chosen as follows with an assumption that knowledge of the
second side reaction ($3B \rightleftharpoons C $) is not available
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - \dfrac{C_B}{\tau}
\end{align*}
The control input is $u = C_{Af}$, the states in the plant model are $x = [C_A,
C_B, C_C]'$, and the measurements are $y = [C_A, C_B]'$. We assume the following
parameters values, $k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, \ k_2 = 0.01 \
\textnormal{m}^3/\textnormal{min}, \ k_3 = 0.05 \
\textnormal{m}^3/\textnormal{min}$, and $\tau = 5 \ \textnormal{min}$.

We examine the performance of the hybrid modeling framework
\eqref{eq:hybrid_discrete_time} in simulation studies. First, we generate
training data by simulating the plant model using a pseudo-random-binary-signal
(PRBS) manipulated input profile. Second, we train a black-box neural network
and the hybrid model \eqref{eq:hybrid_discrete_time} with network architectures
of [9, 16, 2] and [8, 16, 2]. Third, we examine the trained model predictions
and a mean squared error (MSE) performance metric on validation data generated
using the plant model. Finally, we analyze the steady-state cost curves obtained
using the trained models and their differences with the steady-state cost curve
of the plant model. These simulations and analyses shed light on the data
requirements of the hybrid and black-box modeling frameworks, and enable an
understanding of their performance if used for real-time economic optimization
in chemical plants.

The Figure \ref{fig:validation_pred} shows predictions of the trained black-box
and hybrid models after training with 8 hours (480 samples) of data. We observe
that both the models can match the plant data accurately. The black-box model
shows slightly worse predictions in a few places. However, both the trained
models are adequate for real-time optimization in a feedback control system.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[page=9, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Mean squared error (MSE) performance metric obtained with the
           black-box and hybrid model with increase in the number of training
           samples. We notice that the hybrid model slightly outperforms the
           black-box model in data requirements.}
  \label{fig:validation_mse}
\end{figure}

@@

\begin{align*}
  \underset{u_s}{\textnormal{min}} & \quad \ell(y_s, u_s) \\
  x_s &= f(x_s, u_s)\\ 
  y_s &= h(x_s)
\end{align*}
in which $x_s$, $y_s$, $u_s$ are the state, output, and control respectively at
steady-states. For the nonlinear reactor example, the cost is chosen as
$\ell(y_s, u_s) = c_1C_{Af} - c_2C_B$, which measures the raw material cost and
product profit. The Figure \ref{fig:cost_curves} shows the cost curves of the
plant, grey-box, black-box, and hybrid model obtained with 8 hours of training
data. We first notice from the cost curve of the grey-box model that an
optimization with the grey-box model results in the optimizer recommending a
full supply of the raw material $A$. Since some of the product $B$ reacts to
form an undesired side product $C$ in the plant, this recommendation incurres a
cost as observed from the plant cost curve. An intuitive approach to correct
this type of plant-model mismatch is to augment the grey-box model with
integrating disturbance models and estimate the plant model mismatch from data
at one steady state. This strategy however, cannot recover the plant cost curve
in the entire actuator constraint set as the integrating disturbance models are
often linear. We notice that the cost curve of the hybrid model is almost the
same as the cost curve of the plant model in the entire actuator constraint set.
The cost curve of the black-box model is also shown and we notice that it is not
as closely aligned with the plant as compared to the hybrid model.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[page=7, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Steady-state cost curve for the nonlinear plant, grey-box, black-box,
  residual, and hybrid model after training with 10 hours of data.}
  \label{fig:cost_error_bb}
\end{figure}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[page=8, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Steady-state cost curve for the nonlinear plant, grey-box, black-box,
  residual, and hybrid model after training with 10 hours of data.}
  \label{fig:cost_error_hybrid}
\end{figure}

>>>>

in which $N_p$ is the number of past measurements and $\Delta$ is the sample
time between two consecutive measurements. There exists a function $\phi(\cdot)$
and a number of delay time steps $N_p$, such that the plant state at a given
time can be written as $x_p(k) = \phi(\mathbf{y}_{k-N_p:k},
\mathbf{u}_{k-N_p:k-1})$. For linear plants, this function $\phi(\cdot)$ is
known analytically in terms of the plant dynamic matrices.

@@

\begin{align*}
  x_{gk} &= x_g(k) \\
  k_1 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k-N_p:k-1}, \mathbf{u}_{k-N_p:k-1}, u) \\
  x_g & := x_g + (\Delta/2)k_1 \\
  k_2 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g & := x_g + (\Delta/2)k_2 \\
  k_3 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g & := x_g + \Delta k_3 \\
  k_4 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g(k+\Delta) &= x_{gk} + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) 
\end{align*}
in which the sequence $\mathbf{y}_{k+0.5-N_p:k-0.5}$ is obtained by linear
interpolation.

The following optimization problem is solved to compute the weights in the
neural network
\begin{align*}
\underset{W_i, b_i}{\textnormal{min}} \sum_{k=0}^{N_s-1} 
\dfrac{1}{N_s}(\hat{y}(k) - y(k))^2
\end{align*}
in which $W_i$, $b_i$ are the weights in each hidden layer of the network, $N_s$
is the number of training samples, $\hat{y}$ are the predictions of the
hybrid grey-box + neural network model, $N_s$ is the number of training
samples, and $y$ are the measurements from the plant. Note that this
optimization requires an estimate of the initial grey-box state. We assume that
before the open-loop system identification experiment is performed for data
collection, the plant is being operated at some steady-state such that input to
output responses of the plant and grey-box model can be aligned with the use of an
integrating disturbance model. We can use the grey-box state at that steady
state as the initial condition for the above multi-step-ahead prediction-error
optimization.

\subsection{Nonlinear reactor example}
We now demonstrate the application of the above hybrid modeling framework
via a simulation study on a simple chemical process example. We consider a nonlinear reactor with two reactions ($A \rightarrow B $ and  $3B
\rightleftharpoons C$), simulated using the following ordinary differential
equations (ODEs)
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - 3k_2C^3_B + 3k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C^3_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}

The grey-box model is chosen as follows with an assumption that knowledge of the
second side reaction ($3B \rightleftharpoons C $) is not available
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - \dfrac{C_B}{\tau}
\end{align*}
The control input is $u = C_{Af}$, the states in the plant model are $x = [C_A,
C_B, C_C]'$, and the measurements are $y = [C_A, C_B]'$. We assume the following
parameter values, $k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, \ k_2 = 0.01 \
\textnormal{m}^3/\textnormal{min}, \ k_3 = 0.05 \
\textnormal{m}^3/\textnormal{min}$, and $\tau = 5 \ \textnormal{min}$.

We examine the performance of the hybrid modeling framework
\eqref{eq:hybrid_discrete_time} in simulation studies. First, we generate
training data by simulating the plant model using a pseudo-random-binary-signal
(PRBS) of the manipulated control input. Second, we train a black-box neural network
and the hybrid model \eqref{eq:hybrid_discrete_time} with network architectures
of [9, 16, 2] and [8, 16, 2]. Third, we examine the trained model predictions
and a mean squared error (MSE) performance metric on validation data generated
using the plant model. Finally, we analyze the steady-state cost curves obtained
using the trained models and their differences with the steady-state cost curve
of the plant model. These simulations and analyses shed light on the data
requirements of the hybrid and black-box modeling frameworks, and enable an
understanding of their performance if used for real-time economic optimization
purposes in chemical plants.

The Figure \ref{fig:validation_pred} shows predictions of the trained black-box
and hybrid models after training with 8 hours (480 samples) of data. We observe
that both the models can match the plant data accurately. The black-box model
shows slightly worse predictions in a few places. However, both the trained
models are adequate for real-time optimization purpose in a feedback control system.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[page=9, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Mean squared error (MSE) performance metric obtained with the
           black-box and hybrid model with increase in the number of training
           samples. We notice that the hybrid model slightly outperforms the
           black-box model in data requirements and 
           overall prediction accuracy.}
  \label{fig:validation_mse}
\end{figure}

@@

\begin{align*}
  \underset{u_s}{\textnormal{min}} & \quad \ell(y_s, u_s) \\
  x_s &= f(x_s, u_s)\\ 
  y_s &= h(x_s) \\
  \underline{u} \leq &u_s \leq \overline{u}
\end{align*}
in which $x_s$, $y_s$, $u_s$ are the state, output, and control respectively at
steady-states. For the nonlinear reactor example, the cost is chosen as
$\ell(y_s, u_s) = c_1C_{Af} - c_2C_B$, which measures the raw material cost and
product profit. The Figure \ref{fig:cost_curves} shows the cost curves of the
plant, grey-box, black-box, and hybrid model obtained with 8 hours of training
data. We first notice from the cost curve of the grey-box model that an
optimization with the grey-box model results in the optimizer recommending a
full supply of the raw material $A$. Since some of the product $B$ reacts to
form an undesired side product $C$ in the plant, this recommendation incurs a
cost as observed from the plant cost curve. An intuitive approach to correct
this type of plant-model mismatch is to augment the grey-box model with
integrating disturbance models and estimate the plant model mismatch from data
at one steady state. This strategy however, cannot recover the plant cost curve
in the entire actuator constraint set as the integrating disturbance models are
often linear. We notice that the cost curve of the hybrid model is almost the
same as the cost curve of the plant model in the entire actuator constraint set.
The cost curve of the black-box model is also shown and we notice that it is not
as closely aligned with the plant as compared to the hybrid model.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[page=7, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Variation of the cost error metric ($e(C_{Af})$) in the actuator
           constraint set with increase in the training data for the 
           black-box neural network. The metric improves in the 
           constraint set with an increase in training data.}
  \label{fig:cost_error_bb}
\end{figure}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[page=8, width=\textwidth]{tworeac_plots_nonlin.pdf}
  \caption{Variation of the cost error metric ($e(C_{Af})$) in the actuator
  constraint set with increase in the training data for the 
  hybrid grey-box and neural network model. The metric improves in the 
  constraint set with an increase in training data.}
  \label{fig:cost_error_hybrid}
\end{figure}
